---
title: Sonnet to infix function
author: Dominik Rafacz
date: '2022-08-10'
slug: sonnet-to-infix-function
categories: ["Essay"]
tags: ["essay", "operator", "function", "magrittr", "rlang", "tidysq", "language feature", "NSE"]
---

# Some poetry for the programmers
 
> Blest be the day, and blest be the month and year,
>
> Season and hour and very moment blest,
>
> The lovely IDE where first possessed
>
> By two percent signs I found me prisoner; (...)
>
> ~ Francesco Petrarch, Sonnet 61. Translated by by Joseph Auslander. Possibly with some spicing it up by me 

# Subject and addressee of the poem

Custom infix functions are one of my favorite features in R. This article is my love letter to them. But first, quick recap. 

For those that are unfamiliar with the terminology, *infix function* is a function `fun` which is called using `infix notation`, e.g. `x fun y` instead of `fun(x, y)`. They are also called *infix operators* by base R and I am going to use those terms interchangeably, as well as term *infixes*. There is a lot of infix operators in base R used very frequently, i.e. arithmetic or logical operators. We use them so often that we usually forget that they are functions. And that we can **call them just like regular functions**.

```{r}
23 + 19
`+`(23, 19)
```

As you may have noticed, when calling such function in a non-infix-manner, we need to use backticks around the operator itself to avoid calling it in a regular manner.

To prove that they share a lot of common behavior with functions, I am going to demonstrate you how to **redefine them just like regular functions**. We can, for example, make `+` operator work like a multiplication.

```{r}
`+` <- function(lhs, rhs) lhs * rhs
23 + 19
```

Let it ring out loud: **you should not do that at all**. Overriding default behavior of functions may be really dangerous, as it interferes with other chunks of code that were using the operator. I showed this trick only for demonstration purposes.

(Side note: you might be aware that some packages actually change the default behavior of operators, e.g. `ggplot2` has its own `+` for joining plot elements. However, this is slightly different situation. They do not create whole new function, but only add a *method* for *generic function*. They not *override* the behavior, but more like *overload* it.)

Additional remark to make: in this assignment call backticks are also necessary. Alternative is using quotes, but it is inadvisable and its presence in the language is completely due to the legacy reasons. You can read more about it in [Hadley's book](https://adv-r.hadley.nz/names-values.html#non-syntactic). In [another chapter](https://adv-r.hadley.nz/functions.html?q=infix#infix-functions) you can also find more about infix functions.

```{r include=FALSE}
rm("+")
```

Once again: **do not override built-in infix functions**. BUT. **Create custom infix functions!**

R allows for creating your own operators using double **percent signs**. There are actually a few shipped with the base R packages, including inclusion operator `%in%` (pun intended) or matrix multiplication operator `%*%`. But more can be created. And we will explore it in next section.

# Enumeration is an important stylistic device

## Piping operators

The most widely know example is piping operator: `%>%`. I assume that you are familiar with it. If not, [you better get to know dplyr package](https://dplyr.tidyverse.org/). Pipe itself comes from [magrittr](https://magrittr.tidyverse.org/) package, but dplyr show its real strength and greatest advantage of using custom operators: **custom infixes make code more readable**. When we apply operations one by one it makes more sense to write those operations in order of application, not in the inverse order, as we do without infix operators. The shape of the operator itself is also important, as it suggest the direction of the data flow. For me it is more natural to read and understand

```r
x %>% 
  fun_1 %>% 
  fun_2 %>% 
  fun_3

```

rather than

```r
fun_3(fun_2(fun_1(x)))
```

In fact, it became so popular that R 4.1.0 included its own pipe, `|>`. However, I am not going to dwell more on that, because lots have already been said on pipes by others. I only want to mention, what is often forgotten by newcomers, that magrittr offer more than one type of pipe. Go check them out, if you don't know them!

## NULL-default operator

Another great and simple function is `%||%`. It is exported by *rlang* package (it's the turtleblog after all", so rlang mention is obligatory), but other packages borrow this idea to avoid dependency. The code is really simple and quite clearly explains what the function is intended to do.

```{r}
`%||%` <- function(lhs, rhs) if (is.null(lhs)) rhs else lhs

34 %||% 8
NULL %||% 10
```

This one comes really handy when you have optional `NULL`s returned by function or possibly `NULL` parameters of function or named arguments of function. In the code it looks way clearer than expansion of the function definition itself. I usually do not export such a helper function to the end user. But it is worth repeating that **readability of your code for yourself should be as important as readability for others**. If such infixes help you, then you should definitely use them.

## NULL-propagating operator

Let's stay with `NULL`-related functions for a little longer. Now it is time for my own creation. When I build packages or shiny apps it happens quite often that I need to apply some function to some object, but if it is `NULL` I simply need to return `NULL`. For those situations I created `%?>%` operator:

```{r}
`%?>%` <- function(lhs, rhs) if (is.null(lhs)) NULL else rhs(lhs)

7 %?>% exp
NULL %?>% exp
```

This is actually some modification of a regular pipe, but serves a specific purpose. It simplified my code significantly and I love it for its **conciseness**.

## Motif inclusion operator

Now, example from one of our packages, *tidysq*. This is a package for tidy processing of biological sequences. Here we implemented `%has%` operator that checks for presence of specific motif in sequences.

```{r message=FALSE}
library(tidysq)

sq(c("AAAA", "AGCA", "CGCG", "TTCG")) %has% "GC"

```

Me and Laura found it very in line with tidyverse philosophy of code being readable and understandable by others. This can be especially seen in `dplyr` processing pipes.

```{r message=FALSE}
library(dplyr)

tibble(
  id = 1:4,
  sequence = c("LVGWEK", "KLLCVN", "ER", "LLLY")
 ) %>%
  mutate(sequence = sq(sequence)) %>%
  filter(sequence %has% "LL")

```

## Ternary if operator

Finally, one fancy trick! Have you programmed in any language that contains ternary if operator? E.g. in C++ it looks like this:

```c++
(1 > 0) ? 'A' : 'B'

// would return 'A'
```

If expression on lhs of `?` is TRUE, value between `?` and `:` is returned. Otherwise, value after `:` is returned. Why using it over regular if-else? It is an *expression*, while regular `if` in C and C++ is not an expression. Thus, result of this expression can be assigned to any variable.

Do you miss it in R? Probably not. In R `if` is an expression, so we have:

```{r}
if (1 > 0) "A" else "B"
```

But what if we *really* wanted to have something in R more like in C++? Here you go!

```{r}
`%?%` <- function(lhs, rhs) {
  values <- rlang::enexpr(rhs)
  if (values[[1]] != as.symbol(":"))
    stop("RHS for `%?%` operator has to be in form of '*:*' where '*' are any expressions")
  if (lhs) rlang::eval_bare(values[[2]]) else rlang::eval_bare(values[[3]])
}

(1 > 0) %?% "A" : "B"
```

It did it only because *I can*. 

As you can see, I also abused the `:` operator. Not overrode, *abused*. Inside the function rhs argument is in form `expr:expr`. Normally that would result in range. But here I used the **non-standard evaluation**, using `rlang::enexpr()` and `ralng::eval_bare()` to actually stop R from doing what it usually does. I will probably go into detail about it some other time, right now enjoy the fanciness.

## Remark about precedence

One thing to remember: **operators have specific precedence**. Those that have been programming for longer know that well from autopsy, those that have just started programming *will* get to know that. Custom infix functions obey those rules as well and you sometimes may get surprised. You can check that operators provided by me may sometimes be misleading and require parentheses to work properly. It is worth to check out [R operator precedence manual page](https://rdrr.io/r/base/Syntax.html) when in doubt.

# The strange sibling of my love

As a side note I wanted to mention one operator I was not aware of before writing the article. Or: I was not fully aware what lies underneath it.

Have you came across `:=` operator in R? Simply `:=`, with no `%` around it. If you are using data.table or some rlang magic then probably you have seen it. So have I. However, I took it for granted that the reason for its operator existence is due to non-standard evaluation, since it is always used within some specific context (similarly to my previously shown `%?%` operator). I was astonished to find out that I can actually define this

You can write something like this.

```{r}
`:=` <- function(lhs, rhs) lhs + 5 * rhs
2 := 2
```

And *it works*. 

I told you earlier that overriding built-in operators is a *big no-no*. What is the difference with this one. Well, **there is no such operator in base R!** It turns out that `:=` is reserved due to legacy reasons and parser still treats it as a single operator. (You can try creating similar functions for other strings -- you won't make R treat them as infix operators without percent signs, they need to be reserved.) Yet `:=` has no definition in base. You can read more on [the SO thread about this operator](https://stackoverflow.com/questions/32817780/what-is-the-r-assignment-operator-for).

If you want to be fancy, you can assign classic assignment operator to this one and make others wonder how.

```{r}
`:=` <- `<-`
a := 3
```

(But seriously: don't do it.)

# Send the letter!

The ability to create your own inter-argument operators is a very elegant addition to the language. They can make the code much more readable, and in some cases definitely shorter. I strongly encourage you to take advantage of the language's possibilities and play around with it. Since it's your working tool, let's make using it as enjoyable as possible!
